---
title: "Data visualization II"
author: ""
date: ""
output: html_document
editor_options: 
  chunk_output_type: console
---

## Main ideas

- There are different types of variables. 
- Visualizations and summaries of variables must be consistent with the 
  variable type.

## Lecture notes and exercises

Run the following code in the **console** to configure `git`. Fill in your 
name and the email address associated with your GitHub account.

```{r eval = FALSE}
library(usethis)
use_git_config(user.name= "your name", user.email="your email")
```

Next load the `tidyverse` package. 

```{r message = FALSE}
library(tidyverse)
```

There are two types of variables **numeric** and **categorical**.

### Types of variables

Numerical variables can be classified as either **continuous** or **discrete**. 
Continuous numeric variables have an infinite number of values between any 
two values. Discrete numeric variables have a countable number of values.

- height
- number of siblings

Categorical variables can be classified as either **nominal** or **ordinal**. 
Ordinal variables have a natural ordering.

- hair color
- education

### Numeric variables

To describe the distribution of a numeric we will use the properties below.

- shape
  - skewness: right-skewed, left-skewed, symmetric
  - modality: unimodal, bimodal, multimodal, uniform
- center: mean (`mean()`), median (`median()`)
- spread: range (`range()`), standard deviation (`sd()`), interquartile range (`IQR()`)
- outliers: observations outside the pattern of the data

We will continue our investigation of home prices in Minneapolis, Minnesota.

```{r load-data, message = FALSE}
mn_homes <- read_csv("data/mn_homes.csv")
```

Add a `glimpse()` to the code chunk below and identify the following variables 
as numeric continuous, numeric discrete, categorical ordinal, or categorical
nominal. 

- **area**: numeric continuous
- **beds**: numeric discrete
- **community**: categorical

```{r glimpse-data}
glimpse(mn_homes)
```

We can use a **histogram** to summarize a single numeric variable.

```{r histogram}
ggplot(data = mn_homes, 
       mapping = aes(x = salesprice)) + 
   geom_histogram(bins = 25, fill = "lightblue", color = "grey")
```

A **density plot** is another option. We just connect the boxes in a histogram
with a smooth curve.

```{r density-plot}
ggplot(data = mn_homes, 
       mapping = aes(x = salesprice)) + 
   geom_density(fill = "darkgreen", alpha = 0.5) +
  theme_minimal()
```

Side-by-side **boxplots** are helpful to visualize the distribution of a numeric
variable across the levels of a categorical variable.

```{r boxplots}
ggplot(data = mn_homes, mapping = aes(x = community, y = salesprice)) + 
  geom_boxplot(color = "darkviolet", fill = "grey60") + 
  coord_flip() +
  theme_bw()

```

**Question:** What is `coord_flip()` doing in the code chunk above? Try removing
it to see.

Function `coord_flip()` transposes the `x` and `y` axes. Be careful with
`labs()`.

### Categorical variables

**Bar plots** allow us to visualize categorical variables.

```{r bar-plot}
ggplot(data = mn_homes) + 
  geom_bar(mapping = aes(x = reorder(community, community, length))) + 
  coord_flip() +
  labs(x = "Community", y = "Count") +
  theme_light(base_size = 20)
```

**Segmented bar plots** can be used to visualize two categorical variables.

```{r segmented-bar-plot}
ggplot(data = mn_homes, mapping = aes(x = community, fill = fireplace)) + 
  geom_bar() +
  scale_fill_manual(values = c("purple", "orange")) +
  coord_flip()
```

```{r segmented-bar-plot-fill}
ggplot(data = mn_homes, mapping = aes(x = community, fill = fireplace)) + 
  geom_bar(position = "fill") + 
  scale_fill_manual(values = c("purple", "orange")) +
  coord_flip()
```

**Question:** Which of the above two visualizations should be preferred?

The first plot is helpful for within community comparison of having a fireplace
or not. The second plot makes it easy for between community comparison of
having a fireplace or not.


There is something wrong with each of the plots below. Run the code for each 
plot, read the error, then identify and fix the problem.

```{r fix-errors, eval = FALSE}
ggplot(data = mn_homes) + 
  geom_point(mapping = aes(x = lotsize, y = salesprice,
                           shape = 21, size = .85))

ggplot(data = mn_homes) + 
  geom_point(x = lotsize, y = area, shape = 21, size = .85)

ggplot(data = mn_homes) +
  geom_point(mapping = aes(x = lotsize, y = area),
             color = community, size = .85)

ggplot(data = mn_homes) +
  geom_point(mapping = aes(x = 1otsize, y = area))
```

```{r solutions, eval = FALSE}
ggplot(data = mn_homes) + 
  geom_point(mapping = aes(x = lotsize, y = salesprice),
                           shape = 21, size = .85)

ggplot(data = mn_homes) + 
  geom_point(aes(x = lotsize, y = area), shape = 21, size = .85)

ggplot(data = mn_homes) +
  geom_point(mapping = aes(x = lotsize, y = area, color = community), size = .85)

ggplot(data = mn_homes) +
  geom_point(mapping = aes(x = lotsize, y = area))
```

General principles for effective data visualization

- keep it simple
- use color effectively
- tell a story

Take a look at https://github.com/GraphicsPrinciples/CheatSheet/blob/master/NVSCheatSheet.pdf
for how to think through creating an effective visualization.

## Practice

(1) Modify the code outline to create a faceted histogram examining the 
    distribution of year built within each community. When you are finished, 
    remove `eval = FALSE` and knit the file to see the changes.

```{r eval = FALSE}
ggplot(data = mn_homes, mapping = aes(x = yearbuilt)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(~community) +
  labs(x = "Year", y = "Count", title = "Home ages in Minneapolis, MN", 
      subtitle = "Faceted by community")
```

## Additional Resources

- https://raw.githubusercontent.com/rstudio/cheatsheets/master/data-visualization-2.1.pdf
- https://github.com/GraphicsPrinciples/CheatSheet/blob/master/NVSCheatSheet.pdf
- https://ggplot2.tidyverse.org/
- http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html
- https://medium.com/bbc-visual-and-data-journalism/how-the-bbc-visual-and-data-journalism-team-works-with-graphics-in-r-ed0b35693535
- https://ggplot2-book.org/
- https://ggplot2.tidyverse.org/reference/geom_histogram.html
- https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf